Parsing Postfix log files

### 2007-09-24 doc/logparser.tex
Expanded the additional complications section, explaining where in the algorithm the checks are made, for those complications which didn't already have that information. Minor cleanups.

% Useful stuff for math mode.
\usepackage{amstext}
% Wrapping of URLs.  This is used in the bibliography instead of \url{}.
\usepackage{url}
%\DeclareUrlCommand\urlwrap{}
%\newcommand{\urlwrapfill}[1]{\urlwrap{#1}\hfill}
\newcommand{\parsername}{ASO }
% Include images
\usepackage[final]{graphicx}
% Change how nested enumerate environments are labelled.
\renewcommand{\labelenumii}{\roman{enumii}:}
\renewcommand{\refname}{Bibliography}
% Add the bibliography into the table of contents.
\usepackage[section,numbib]{tocbibind}
% When creating a PDF make the table of contents into links to the pages
% (without horrible red borders) and include bookmarks.  The title and
% author don't work - I think either gnuplot or graphviz clobbers it.
\usepackage{hyperref}
\hypersetup{
    pdftitle    = {Parsing Postfix log files},
    pdfauthor   = {John Tobin},
    final       = true,
    pdfborder   = {0 0 0},
%    breaklinks  = true,
}
\usepackage{breakurl}

@@

Postfix uses several restriction lists (each containing zero or more
restrictions) to decide the fate of an SMTP command, one list for each
stage of the SMTP conversation: client connection, HELO command, MAIL FROM
command, RCPT TO command (possibly multiple times), DATA command, and end
of data.  By default restriction lists will not be evaluated until the
first RCPT TO command is received, because some clients don't deal with
earlier rejections properly; a side benefit of this delay is that Postfix
has more information available when logging the rejection.\footnote{The
\texttt{smtpd\_delay\_reject} parameter controls this behaviour.}  Postfix
uses simple lookup tables as the deciding factor for some restrictions,
e.g.\ \texttt{check\_client\_access~cidr:/etc/postfix/client\_access}
\footnote{\texttt{cidr} is the type of the lookup table \cite{cidr_table}}
will check whether the IP address of the connected client matches the left
hand side of each line in the file and will return the right hand side of
the first matching line.  Other restrictions determine their result in
different ways, e.g.\ \texttt{reject\_rbl\_client rbl.example.com} checks
whether the IP address of the client is present in the listed RBL\@.

@@

\parsername is bundled with extensive documentation, including an in-depth
explanation of the database schema.  It supports multiple values for every
field for which multiple values are reasonable (e.g. the sender address
should have one value only, whereas the recipient address can have
multiple values).  The default database it uses is SQLite3, which supports
most standard SQL features (its lacks full foreign key support), has
libraries available for most common programming languages, and a simple
text based client allowing ad-hoc queries with minimal effort.

Parsing in LMA is hand-coded and requires major changes to the code to
parse new log lines or extract additional data.  It does not appear to deal
with any of the complications discussed in
sections~\ref{complications}~and~\ref{additional complications}.  The
program fails to parse Postfix 2.2.x or 2.3.x log files; indeed it produces
a number of warnings and no usable output.  Parsing in \parsername is
divided in two (see sections~\ref{rules} and ~\ref{parsing-algorithm} for a
more detailed explanation): parsing a new log line is a trivial matter of
adding a new rule, but code changes may be required if new complications
(sections~\ref{complications} and~\ref{additional complications}) must be
dealt with or a new action is required (section~\ref{actions-in-detail}).

LMA does provide some simple reports (IP-STORY, FROM-STORY, DAILY-EMAIL and
DAILY-REJECT); the author was unable to parse any logs with LMA, so didn't
have an opportunity to review the reports.  \parsername does not have any
bundled reports, it delegates that task to a separate, dedicated program.

@@

XXX EXPLAIN EXACTLY WHERE IN THE ALGORITHM THE \newline CHECKS ETC\@. KICK
IN\@.

@@

@@

@@

These checks are made during the DISCONNECT action: if all checks are
successful then the mail is assumed to have been discarded when the client
disconnected.  There will be no further entries logged for such mails, so
without identifying and discarding them they accumulate in the state table
and will cause clashes if the queueid is reused.  The mail cannot be
entered in the database as the only data available is the client hostname
and IP address, but the database schema (see section~\ref{database schema})
requires many more fields be populated.

@@

@@

@@

@@

Occasionally a log file will have out of order log lines which cannot be
dealt with by the techniques described in sections~\ref{tracking
re-injected mail},~\ref{discarding cleanup lines}~and~\ref{pickup logging
after cleanup}.  In the 93 log files used for testing this occurs only five
times in 607,21,709 log lines, but for completeness of the algorithm it
should be dealt with.  The five occurrences in the test log files have the
same characteristics: the \texttt{postfix/local} log line showing delivery
to a local mailbox occurs after the \texttt{postfix/qmgr} log line showing
removal of the mail from the queue because delivery is completed.  This
causes problems: the mail is not complete, so entry into the database
fails; a new mail is created when the \texttt{postfix/local} line is parsed
and remains in the state tables; four warnings are issued per pair of out
of order log lines.

The solution to this problem is to examine the list of programs which have
logged messages for each mail, comparing the list against a table of
known-good combinations of programs.  If the mail's combination is found in
the valid list the mail can be entered in the database; if the combination
is not found entry must be postponed and the mail flagged.  The
SAVE\_BY\_QUEUEID action checks for the flag and retries entry if it's
found; if the additional log lines have caused the mail to reach a valid
combination entry will succeed, otherwise it must be postponed once more.

@@

\input
\input{build/stats-normal-filesize-line-count-include}

>>>>

% Useful stuff for math mode.
\usepackage{amstext}
% Wrapping of URLs.  This is used in the bibliography instead of \url{}.
\usepackage{url}
%\DeclareUrlCommand\urlwrap{}
%\newcommand{\urlwrapfill}[1]{\urlwrap{#1}\hfill}
\newcommand{\parsername}{ASO}
% Include images
\usepackage[final]{graphicx}
% Change how nested enumerate environments are labelled.
\renewcommand{\labelenumii}{\roman{enumii}:}
\renewcommand{\refname}{Bibliography}
% Add the bibliography into the table of contents.
\usepackage[section,numbib]{tocbibind}
% When creating a PDF make the table of contents into links to the pages
% (without horrible red borders) and include bookmarks.  The title and
% author don't work - I think either gnuplot or graphviz clobbers it.
\usepackage{hyperref}
\hypersetup{
    pdftitle    = {Parsing Postfix log files},
    pdfauthor   = {John Tobin},
    final       = true,
    pdfborder   = {0 0 0},
%    breaklinks  = true,
}
\usepackage{breakurl}

@@

Postfix uses several restriction lists (each containing zero or more
restrictions) to decide the fate of an SMTP command, one list for each
stage of the SMTP conversation: client connection, HELO command, MAIL FROM
command, RCPT TO command (possibly multiple times), DATA command, and end
of data.  By default restriction lists will not be evaluated until the
first RCPT TO command is received, because some clients don't deal with
earlier rejections properly; a side benefit of this delay is that Postfix
has more information available when logging the rejection.\footnote{The
\texttt{smtpd\_delay\_reject} parameter controls this behaviour.}  Postfix
uses simple lookup tables as the deciding factor for some restrictions,
e.g.\ \texttt{check\_client\_access~cidr:/etc/postfix/client\_access}
\footnote{\texttt{cidr} is the type of the lookup table~\cite{cidr_table}}
will check whether the IP address of the connected client matches the left
hand side of each line in the file and will return the right hand side of
the first matching line.  Other restrictions determine their result in
different ways, e.g.\ \texttt{reject\_rbl\_client rbl.example.com} checks
whether the IP address of the client is present in the listed RBL\@.

@@

\parsername{} is bundled with extensive documentation, including an in-depth
explanation of the database schema.  It supports multiple values for every
field for which multiple values are reasonable (e.g.\ the sender address
should have one value only, whereas the recipient address can have
multiple values).  The default database it uses is SQLite3, which supports
most standard SQL features (its lacks full foreign key support), has
libraries available for most common programming languages, and a simple
text based client allowing ad-hoc queries with minimal effort.

Parsing in LMA is hand-coded and requires major changes to the code to
parse new log lines or extract additional data.  It does not appear to deal
with any of the complications discussed in
sections~\ref{complications}~and~\ref{additional complications}.  The
program fails to parse Postfix 2.2.x or 2.3.x log files; indeed it produces
a number of warnings and no usable output.  Parsing in \parsername{} is
divided in two (see sections~\ref{rules} and~\ref{parsing-algorithm} for a
more detailed explanation): parsing a new log line is a trivial matter of
adding a new rule, but code changes may be required if new complications
(sections~\ref{complications} and~\ref{additional complications}) must be
dealt with or a new action is required (section~\ref{actions-in-detail}).

LMA does provide some simple reports (IP-STORY, FROM-STORY, DAILY-EMAIL and
DAILY-REJECT); the author was unable to parse any logs with LMA, so didn't
have an opportunity to review the reports.  \parsername{} does not have any
bundled reports, it delegates that task to a separate, dedicated program.

@@

@@

This check is performed during the COMMIT action.

@@

This check is performed in the DISCONNECT action; it requires support in
the CLONE action where a list of cloned connections is maintained.

@@

These checks are made during the DISCONNECT action: if all checks are
successful then the mail is assumed to have been discarded when the client
disconnected.  There will be no further entries logged for such mails, so
without identifying and entering them in the database immediately they
accumulate in the state table and will cause clashes if the queueid is
reused.

@@

During the TIMEOUT action if the mail lacks a cleanup line, the queueid and
other data about the mail, particularly the timestamp of the last log line,
are added to a global cache of timeout queueids.  This cache is
subsequently consulted during the MAIL\_PICKED\_FOR\_DELIVERY action, and
if the queueid exists and the cleanup line passes the checks listed above
it will be discarded.

@@

This complication is dealt with during the PICKUP action.

@@

These occurrences are handled by the three actions POSTFIX\_RELOAD,
SMTPD\_DIED and SMTPD\_WATCHDOG\@.

@@

Occasionally a log file will have out of order log lines which cannot be
dealt with by the techniques described in sections~\ref{tracking
re-injected mail},~\ref{discarding cleanup lines}~and~\ref{pickup logging
after cleanup}.  In the 93 log files used for testing this occurs only five
times in 60,721,709 log lines, but for completeness of the algorithm it
should be dealt with.  The five occurrences in the test log files have the
same characteristics: the \texttt{postfix/local} log line showing delivery
to a local mailbox occurs after the \texttt{postfix/qmgr} log line showing
removal of the mail from the queue because delivery is completed.  This
causes problems: the mail is not complete, so entry into the database
fails; a new mail is created when the \texttt{postfix/local} line is parsed
and remains in the state tables; four warnings are issued per pair of out
of order log lines.

The solution to this problem is to examine the list of programs which have
logged messages for each mail, comparing the list against a table of
known-good combinations of programs (this check is performed during the
COMMIT action).  If the mail's combination is found in the valid list the
mail can be entered in the database; if the combination is not found entry
must be postponed and the mail flagged.  The SAVE\_BY\_QUEUEID action
checks for the flag and retries entry if it's found; if the additional log
lines have caused the mail to reach a valid combination entry will succeed,
otherwise it must be postponed once more.

@@

\input{build/stats-normal-filesize-line-count-include}
